---
title: "Gov 50 Final Project"
author: "Ariana Figueroa"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction

I am interested in exploring data related to food deserts in urban areas and with a predominantly low-income population. I think that the data I find will most likely reflect that low-income communities in urban spaces have a harder time finding fresh foods, meaning that they live in food deserts. Throughout this project, I will try to address the question; does urbanization and low-income rates among counties in California make certain communities more susceptible to food deserts? I believe that wealthier communities will have lower food desert rates in comparison to low-income communities, and that those living in urban spaces have less access to fresh foods due to the density of cities and convenience of fast food chains. 

```{r echo= FALSE}
library(tidyverse)
library(readxl)
library(dplyr)
food_access <- read_excel("~/GOV 50/PSETS/gov50-final-project/FoodAccessResearchAtlasData2019.xlsx", 
                          sheet = 3, 
                          na = c("", "NA", "NULL")) |> 
mutate(urban = if_else(Urban == 1, "Urban", "Rural"))

```

## Data Source

The data set I used in this research study is pulled from the Data.gov Data sets released by the government titled, "Food Access Research," a survey index collected through the Census.The data set collected information from households across the nation, and I decided to focus my study on counties in California. The key dependent factor is the population of People Living in Food Deserts. The independent factors of this comparative analysis are the Median Poverty Rates per county and Urbanization (whether someone lives in a rural town, suburbs, or city). I plan to identify and analyze whether or not these independent factors impact peoples' susceptibility of experiencing food deserts, which I hypothesize does have an impact.

I base this hypothesis on the fact that inner city/urban communities lack the spaces for supermarkets and there is prior data exploring the reason why supermarkets are so hard to access in urban areas where there are usually more low-income, people of color inhabiting the area. The unit of analysis I will be basing my observation on is population ("x" amount of people living in food deserts).The unit of analysis for my independent variables will be average income per county in dollars (per California Counties), and urbanization. The explanatory variables of interest are urbanization and income. My outcome variable of interest will be whether or not income and/or urbanization have an impact on the probability of communities living in food deserts. If I observe that those counties that are considered to be more high-income have more supermarkets with closer proximity than low-income communities, my hypothesis will be supported. If the data happens to show no correlation between incomes and Urbanization, then my hypothesis will disproved. 



```{r echo = FALSE}
califonia_counties <- food_access |>  
  filter(State == "California") |>
    select(- c(lakidshalf:laomultirhalfshare,
            lakids20:laomultir20share,
            lakids1:lapop10share,
            lahunv20:lasnap20share,
            TractKids:TractSNAP)) |>
  mutate(urban = if_else(Urban == 1, "Urban", "Rural"))
  
food_access$lahisphalf <- as.numeric(as.character(food_access$lahisphalf))


```


```{r echo = FALSE}
california_counties <- food_access|>
  filter(State == "California")

california_counties$MedianFamilyIncome <- as.numeric(as.character(california_counties$MedianFamilyIncome))
california_counties$LALOWI05_10 <- as.numeric(as.character(california_counties$LALOWI05_10)) 


california_counties <- california_counties |>
  group_by(County) |>
  summarize(county_pop = sum(Pop2010, na.rm = TRUE),
            county_avg_income = mean(MedianFamilyIncome, na.rm = TRUE),
            avg_pop_fd = mean(LALOWI05_10, na.rm = TRUE),
            hisp_pop = sum(lahisphalf, na.rm = TRUE),
            urban = sum(Urban, na.rm = TRUE),
            med_poverty_rate = median(PovertyRate, na.rm = TRUE)) 


```

 
```{r echo = FALSE}
library(ggplot2)
ggplot(data = california_counties,
       mapping = aes(x = county_avg_income,
                     y = avg_pop_fd)) +
  
  geom_point() + 
  labs(
    title = "People Living in Food Deserts",
    x = "Average Income Across Counties",
    y = "Population of People") + geom_smooth()
  

```
The graph shows the relationship between income and food deserts. The Y variable is the low income population count beyond 1/2 mile for urban areas or 10 miles for rural areas from supermarket. This means that approximately an average 1500 people who make about $50,000 a year, are not of close proximity to supermarkets. Of those who are in the higher income bracket,there is a smaller number of individuals are far from supermarkets. The line of best fit shows a slightly negative relationship concerning exposure to food deserts and higher income. As people are wealthier, they are less likely to experience living in food deserts.  

This plot focuses on one of my independent variables, which is income and it highlights the disparities between people across counties, not only by income, but by exposure to food deserts as well. This correlation is interesting as it helps prove my hypothesis that wealthier counties have a smaller population experiencing food deserts. Low-income communities are at a disadvantage as the highest report lands above the federal poverty line. 


```{r echo = FALSE}
counties_urbanization<- california_counties|>
  mutate(urbanization = if_else(urban > 25, "Urban", "Rural" )) |>
  select(-c(urban))


try_1<- counties_urbanization |>
  group_by(urbanization) |>
  pivot_wider(
    names_from = urbanization,
    values_from = avg_pop_fd)

```


## Food Deserts in Urban vs. Rural Counties


```{r echo = FALSE}
ggplot(data = counties_urbanization,
       mapping = aes(x = avg_pop_fd, fill = urbanization)) + geom_histogram(position = "identity", alpha = 0.6, bins = 10) + 
  labs(
    title = "Food Deserts in Urban vs. Rural Counties",
    x = "Average Population of People Living in Food Deserts",
    y = "County Count"
  )
```

This plot highlights the fact that most people living in Food Deserts come from Urban spaces. However, it is interesting that the county with the highest reported population of people comes from a rural county. This can be explained by the fact that rural communities often have lower density rates, meaning that many families have long drives to get to supermarkets or other spaces where they can access fresh foods. Another possibility could be the fact that these are agriculture communities, dismissing the need for supermarkets, yet still being far from them. Despite this outlier in the data, it is safe to say that most people living in food deserts consistently come from urban spaces. As mentioned prior, many studies have shown that the lack of space diminishes the opportunity to establish a farmers market or develop a permanent food market for families. 


## Linear Regression

```{r echo = FALSE}
library(broom)
library(tidyverse)


pov_prox <- lm(avg_pop_fd ~ med_poverty_rate + urbanization, data = counties_urbanization) %>%
  broom::tidy() %>%
  select(term, estimate, std.error, p.value)


pov_prox$term <- ifelse(pov_prox$term == "(Intercept)", "Constant",
                        ifelse(pov_prox$term == "med_poverty_rate", "Median Poverty Rate",
                               ifelse(pov_prox$term == "urbanizationUrban", "Urbanization", pov_prox$term)))


knitr::kable(pov_prox, digits = 3, col.names = c("Variable", "Coefficient Estimate", "Standard Error", "P-Value"))
```
The regression analysis suggests that there is a relationship between Median Poverty Rate across California counties and the average number of people living in food deserts. For each unit increase in the median poverty rate, the average population living in food deserts, in this case presented as the variable "intercept", increases by 40 people, assuming all other factors remain constant. The Median Poverty Rate per County is significant as it has a P-Value of 0, so the null hypothesis is rejected, meaning that living in poverty does have an impact on whether or not you are more or less likely to live in food deserts. The Urbanization variable is also statistically significant with a P-Value of 0.046, meaning that we can reject the null hypothesis for this variable as well. Considering the result of the regression, it is valid to say that there is a correlation between food deserts and whether or not people are living in urban or rural spaces.  


## Conclusion 

The results and assessments of poverty and urbanization's impact on food access proves to be significant. Both graphs and the linear regression show poverty rates and urbanization to be a factor indicating whether or not a population is more susceptible to experiencing food deserts, thus supporting my hypothesis. Some possible limitations of my analysis is the possibility for lack of information, since this is a survey I run into the risk of excluding the experiences of some people. A possible confounder can be the communities in which people live in and how some rural communities in California are technically far from supermarkets but have access to fresh foods through neighboring farms. Although the latter is less of a strong confounder; the possible exclusion of people in counties is what leaves me dissatisfied. In order to fix this, I would extend the study and implement canvassers and strive for a response per household. In order to have a more representative analysis, I would also research how food access varies by ehtnicity and education across California counties. 

